%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% This file is part of the book
%%
%% Algorithmic Graph Theory
%% http://code.google.com/p/graph-theory-algorithms-book/
%%
%% Copyright (C) 2009, 2010, 2011 Minh Van Nguyen <nguyenminh2@gmail.com>
%%
%% See the file COPYING for copying conditions.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{Acknowledgments}

\begin{itemize}
\item Fidel Barrera-Cruz: reported typos in
  Chapter~\ref{chap:trees_forests}. See changeset~101. Suggested
  making a note about disregarding the direction of edges in
  undirected graphs. See changeset~277.

\item Daniel Black: reported a typo in
  Chapter~\ref{chap:introduction}. See changeset~61.

\item Kevin Brintnall: reported typos in the definition of
  $\iadj(v) \cap \oadj(v)$; see changesets~240 and~242. Solution to
  Example~\ref{eg:introduction:walks_paths_trails}(2); see changeset~246.

\item Aaron Dutle: reported a typo in
  Figure~\ref{fig:introduction:bipartite_complete_bipartite_graphs}. See
  changeset~125.

\item P\'eter L. Erd\H{o}s (\url{http://www.renyi.hu/~elp}) for informing
  us of the reference~\cite{ErdosEtAl2010} on the Havel-Hakimi
  theorem for directed graphs.

\item Noel Markham: reported a typo in
  Algorithm~\ref{alg:graph_algorithms:dijkstra_general}. See
  changeset~131 and
  \href{http://code.google.com/p/graph-theory-algorithms-book/issues/detail?id=2}{Issue~2}.

\item Caroline Melles: clarify definitions of various graph types
  (weighted graphs, multigraphs, and weighted multigraphs); clarify
  definitions of degree, isolated vertices, and pendant and using the
  butterfly graph with $5$ vertices~(see
  Figure~\ref{fig:introduction:butterfly_graph}) to
  illustrate these definitions; clarify definitions of trails, closed
  paths, and cycles; see changeset~448. Some rearrangements of
  materials in Chapter~\ref{chap:introduction} to make the reading
  flow better and a few additions of missing definitions; see
  changeset~584. Clarifications about unweighted and weighted
  degree of a vertex in a multigraph; notational convention about a
  graph being simple unless otherwise stated; an example on graph
  minor; see changeset~617.

\item Pravin Paratey: simplify the sentence formation in the definition
  of digraphs; see changeset~714 and
  \href{http://code.google.com/p/graph-theory-algorithms-book/issues/detail?id=7}{Issue~7}.

\item The world map in
  Figure~\ref{fig:graph_algorithms:worldmap_capital_cities} was
  adapted from an SVG image file from Wikipedia. The original SVG file
  was accessed on 2010-10-01 at
  \url{http://en.wikipedia.org/wiki/File:Worldmap_location_NED_50m.svg}.
\end{itemize}
